A Quasi-Genuine FIFO Total Order Multicast Primitive

### 2011-06-21 trunk/textos/optpaxos/draftv3.tex
proof sketch of correctness for uniform integrity

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\cms}{\mbox{QGFTO-Mcast}}
\newcommand{\cmsend}[1]{\mbox{\cms({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{OPT-Deliver}}
\newcommand{\cons}{\mbox{QGFTO-Deliver}}
\newcommand{\rmc}{\mbox{FR-MCast}}
\newcommand{\rmd}{\mbox{FR-Deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcast}[2]{\mbox{\rmc({#1},{#2})}}
\newcommand{\rmdel}[1]{\mbox{\rmd({#1})}}

@@

\blankline
\INDENT{\textit{When} \rmdel{$m'$}}
  \IF{$G = m'.src$}
    \STATE $propPending \leftarrow propPending \cup \{m'\}$
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m'\}$  
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} Decide($k$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$  
    \STATE \rmcast{$m$}{$m.dst \setminus \{G\}$} 
  \ENDINDENT
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$ }}
  \STATE $barPending \leftarrow barPending \setminus \{m\}$
  \STATE \consdel{$m$}
  \STATE $delivered \leftarrow delivered \cup \{m\}$

@@

\blankline
\INDENT{\textit{When} Decide($k$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:chgts}
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$
    \IF{$m \neq null$}
      \STATE \rmcast{$m$}{$(m.dst \setminus \{G\}) \cup blockers(m)$} \label{algline:barreq} \COMMENT{the message is sent to blockers(m), as a barrier request}
    \ELSE
      \STATE \rmcast{$m$}{$m.dst$} \COMMENT{no need to unblock $null$, besides it would create infinite messages}
    \ENDIF
  \ENDINDENT
  \STATE $nextProp \leftarrow k + 1$
  \STATE $k \leftarrow k + 1$
\ENDINDENT

@@

>>>>

\newcommand{\blankline}{\vspace{4 mm}}
\newcommand{\cms}{\mbox{QGFTO-Mcast}}
\newcommand{\cmd}{\mbox{QGFTO-Deliver}}
\newcommand{\cmdel}[1]{\mbox{\cmd({#1})}}
\newcommand{\cmsend}[1]{\mbox{\cms({#1})}}
\newcommand{\tconsm}{T_{cons}}
\newcommand{\tcons}{\mbox{$\tconsm$}}
\newcommand{\opt}{\mbox{OPT-Deliver}}
\newcommand{\cons}{\mbox{QGFTO-Deliver}}
\newcommand{\rmc}{\mbox{FR-MCast}}
\newcommand{\rmd}{\mbox{FR-Deliver}}
\newcommand{\optdel}[1]{\mbox{\opt({#1})}}
\newcommand{\consdel}[1]{\mbox{\cons({#1})}}
\newcommand{\rmcast}[2]{\mbox{\rmc({#1},{#2})}}
\newcommand{\rmdel}[1]{\mbox{\rmd({#1})}}

@@

\blankline
\INDENT{\textit{To} \cms{} \textit{a message $m$}}
  \STATE $m.ts \leftarrow (getTime(),0)$
  %\COMMENT{current wallclock value as the timestamp of $m$}
  \STATE \rmcast{$m$}{$\{G\}$} \label{algline:rmlocal}
%  \INDENT{\textbf{for all} $p' \in m.dst \cup \{G\}$ \textbf{do}}
%    \STATE send($p'$, $m$)    
%    \COMMENT{send optimistically $m$ to all involved processes}
%  \ENDINDENT
\ENDINDENT

\blankline
\INDENT{\textit{When} \rmdel{$m'$}}
  \IF{$G = m'.src$}
    \STATE $propPending \leftarrow propPending \cup \{m'\}$ \label{algline:addtoproppending}
  \ELSIF{$m' \notin barPending \wedge m' \notin delivered$}
    \STATE $barPending \leftarrow barPending \cup \{m'\}$ \label{algline:insbp1}
    \STATE $barrier(m'.src,G) \leftarrow m'.ts$ \label{algline:incbar}
  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in propPending \wedge nextProp = k$}}
%  \STATE $propPending \leftarrow propPending \setminus \{m\}$
%  \IF {$m \notin decided \wedge \nexists m' \in decided: m'.ts > m.ts$}
    \STATE $propPending \leftarrow propPending \setminus decided$%(decided \cup \{m' : \exists m'' \in decided \wedge m'.ts < m''.ts \wedge m' \neq null\})$ 
    \label{algline:removedecided}
    \IF {$propPending \neq \emptyset$}
      \STATE $nextProp \leftarrow k + 1$
      \STATE Propose($k$,$propPending$) \label{algline:propose}
    \ENDIF
%  \ENDIF
\ENDINDENT

\blankline
\INDENT{\textit{When} Decide($k'$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$  \label{algline:insbp2}
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$  \label{algline:addtodecided}
    \STATE \rmcast{$m$}{$m.dst \setminus \{G\}$} \label{algline:rmothers}
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

\blankline
\INDENT{\textit{When $\exists m \in barPending : \forall G' \in sendersTo(G): m.ts < barrier(G',G)$\\ $\wedge\ \nexists m' \in barPending : m'.ts < m.ts$ }} \label{algline:inbarpending}
  \STATE $barPending \leftarrow barPending \setminus \{m\}$
  \STATE \consdel{$m$} \label{algline:consdeliver}
  \STATE $delivered \leftarrow delivered \cup \{m\}$

@@

\blankline
\INDENT{\textit{When} Decide($k'$,$msgSet$)}
%  \STATE $m.gs \leftarrow k$
  \INDENT{\textbf{while} $\exists m \in msgSet : (\forall m' \in msgSet : m \neq m' \Rightarrow m.ts < m'.ts)$ \textbf{do}}
    \STATE $msgSet \leftarrow msgSet \setminus \{m\}$
    \COMMENT{the messages are handled in ascending order of timestamp}
    \IF{$\exists m' \in decided : m'.ts \geq m.ts \wedge (\nexists m'' \in decided : m''.ts > m'.ts)$}
      \STATE $m.ts \leftarrow (m'.ts.rtc, m'.ts.seq + 1)$ \label{algline:chgts}
    \ENDIF     
    \IF{$G \in m.dst$} \label{algline:checkcons}
      \STATE $barPending \leftarrow barPending \cup \{m\}$
    \ENDIF
    \STATE $decided \leftarrow decided \cup \{m\}$
    \IF{$m \neq null$}
      \STATE \rmcast{$m$}{$(m.dst \setminus \{G\}) \cup blockers(m)$} \label{algline:barreq} \COMMENT{the message is sent to blockers(m), as a barrier request}
    \ELSE
      \STATE \rmcast{$m$}{$m.dst$} \COMMENT{no need to unblock $null$, besides it would create infinite messages}
    \ENDIF
  \ENDINDENT
  \STATE $nextProp \leftarrow k' + 1$
  \STATE $k \leftarrow k' + 1$
\ENDINDENT

@@

Here, we prove that the \cms{} primitive ensures the properties defined in section \ref{sec:model} -- uniform validity, uniform agreement, uniform integrity, uniform total order and FIFO order. To do this, we prove that Algorithm \ref{algorithm:deliveryminimal}, when used along with Algorithm \ref{algorithm:nullperiodic}, ensures theses properties.

\subsection{Uniform Validity}

\subsection{Uniform Agreement}

\subsection{Uniform Integrity}

This property means that, if a message $m$ was delivered to process $p$ of group $G$, then it has (1) been multicast before, (2) $G \in m.dst$ and (3) it has not been delivered by $p$ before.

For a message $m$ to be delivered (l. \ref{algline:consdeliver}), it must be contained in the $barPending$ list (from the condition in line \ref{algline:inbarpending}).  There are two possibilities to when $m$ has been inserted into such list:
\begin{itemize}
  \item $m$ has been originated in the same same group of $p$, that is, $p \in m.dst$, which means $m$ was inserted into $barPending$ in line \ref{algline:insbp2} of Algorithm \ref{algorithm:deliveryminimal}, or
  \item $m$ has been sent from a group $G \neq G'$, which means that it was inserted into $barPending$ in line \ref{algline:insbp1}. 
\end{itemize}

In line \ref{algline:insbp2}, $m$ is inserted into $barPending$ only if $G \in m.dst$, which satisfies condition (1), and if it was decided in some consensus instance within $G$. To have been decided within $G$, from the properties of consensus \cite{lamport1998ptp}, we know that it must have been proposed by some process of $G$, which happens in line \ref{algline:propose}. In line \ref{algline:propose}, for a message to be proposed, it must be in the $propPending$, as its contents are the value proposed. For a message to be in the $propPending$ list, it must have been inserted there, which happens only in line \ref{algline:addtoproppending}. For l. \ref{algline:addtoproppending} to be executed, $m$ must have been \rmd{}ed and $G$ must be the source group of $m$, that is, $m.src = G$. For a message to be \rmc{} by a process to its own group, a \cmsend{$m$} call must have been made -- which satisfies condition (2) --, for line \ref{algline:rmlocal} is the only one where a process \rmc{}s a message to its own group.

Also, for a message to be inserted into $propPending$ (l. \ref{algline:addtoproppending}) it cannot have been decided already (l. \ref{algline:removedecided}). We assume here that a process can only make a proposal for consensus instance $k$ when it knows all the values decided in all instances prior to $k$. Therefore, when a process proposes a message, it knows all the messages which were already decided in previous consensus instances. As a decided message is inserted into a $decided$ list (l. \ref{algline:addtodecided}, which is always executed after a decision), a process cannot propose a message which was decided already and, therefore, a message cannot be decided twice (in two different consensus instances). As a message is inserted into $barPending$ by processes of its own group of origin only when decided (l. \ref{algline:insbp2}), it cannot be inserted twice, thus it cannot be delivered twice, satisfying condition (3).

As for processes from destinations groups which are not from the source group of $m$, they will never execute line \ref{algline:insbp2}, since they never decide on $m$. However, processes of the group $m.src$, which is the source of $m$, will \rmc{} $m$ to every $G'$ which is a destination of $m$ other than $G$ (l. \ref{algline:rmothers}). When $G'$ \rmd{}s $m$, it checks whether $m$ was already inserted into $barPending$ or $delivered$, where $delivered$ is the list of all messages which were ever \cmd{}ed. Once a message has been inserted into \mbox{$barPending \cup delivered$}, it stays there forever. Therefore, also messages which arrive from other groups are never inserted twice into $barPending$, so that they are never delivered twice, satisfying (3).

For line \ref{algline:insbp1}, $m$ is inserted into $barPending$ of process $p$ of group $G$ when it has been \rmd{}ed, and only if $m.src \neq G$. For a process $p' \in G' \neq G$ to \rmc{} $m$, $m$ must have been decided within $G'$, which means that it was proposed within $G'$, therefore it was \cms{} -- satisfying condition (1) -- by $G' \neq G$ to $G$, which, from line \ref{algline:rmothers}, necessarily belongs to $m.dst$, satisfying condition (2).

\begin{flushright}
$\square$
\end{flushright}

\subsection{Uniform Total Order}

\subsection{FIFO Order}
